\begin{comment}
  \bibliography{project.bib}
\end{comment}

\chapter{Methods}
\label{cha:method}

\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bfseries +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bfseries +}}

\section{Tools}

\newcommand{\credits}[1]{\footnote{\url{#1}}}

To explore how the aforementioned image formats works, I decided to
write small programs that would analyze the numbers that images in the
formats consists of and dump the raw color. My programming language of
choice for this was \CC{} and my compiler was
\gcc\credits{http://gcc.gnu.org/}. For doing the memory debugging that
is often necessary in \CC{} programming, I used
Valgrind\credits{http://valgrind.org/}, which at times was a real life
saver. My editor of choice for writing the code was
\emacs\credits{http://www.gnu.org/software/emacs/}.

To learn the theory necessary to write these programs I had to read
several books and many articles, all of which are cited in the
bibliography(see the table of contents).

This document was created and typeset using
\LaTeX{}\credits{http://www.latex-project.org/}. I chose to use
\LaTeX{} rather than Word because I, as a programmer, find it a lot
easier to use than Word. The \LaTeX{} source files were also edited in
\emacs, using the
AUC\TeX{}\credits{http://www.gnu.org/software/auctex/} package.

The body text is typeset in the font \TeX{} Gyre Pagella, a clone of
Hermann Zapf's Palatino, and the monospace text uses the font
\texttt{Luxi}.



All the graphics of this document were created using
\tikzname \& \pgf\credits{http://sourceforge.net/projects/pgf/}

All the source files for both this document and the programs were
managed using \git\credits{http://git-scm.com/}, the version control system.

\section{Structure Of This Text}

The chapters in this text are laid out as follows:

\begin{description}
\item[Chapter 4] Any preliminary knowledge that is absolutely
  necessary to understand \textit{any} chapter of this text will be
  put here. We will also establish all the conventions and glossaries
  that we will be using throughout the rest of the text in this
  chapter.

\item[Chapter 5] We will in this chapter discuss how color is
  represented digitally.

\item[Chapter 6] The main topic of this chapter will be the \rle
  compression algorithm. This algorithm is used in the \tga format.

\item[Chapter 7] The \tga format is discussed.

\item[Chapter 8] The \lzw compression algorithm will be explained,
  which is used in the \gif format.

\item[Chapter 9] The \gif format will be the subject of this chapter.

\item[Chapter 10] To ensure the data integrity of \png images something
  known as a \crc is used, which is the topic of this chapter.

\item[Chapter 11] The compression algorithm used in the \png format,
  known as \deflate, is a rather complex algorithm that consists of
  several subalgorithm. One of these subalgorithm is known as
  Huffman Coding, which is what we will be discussing in this chapter.

\item[Chapter 12] The second subalgorithm of the \deflate algorithm is
  known as \lzone, which is the algorithm we will discuss in this
  chapter. An improvement of \lzone called \lzss is also discussed in
  this chapter.

\item[Chapter 13] In this chapter we will combine \lzone and Huffman
  Coding into the final \deflate algorithm.

\item[Chapter 14] In this chapter the \png format will finally be
  discussed.

\item[Chapter 15] All the former chapters will be discussed and we
  will attempt to draw a final conclusion from the project. The
  project as a whole will also be discussed. Answers to the exercises
  can be found at the end of every chapter.

\end{description}

Small exercises can also be found in the text. They were primarily
added so that the reader could have a way of checking that he/she
truly has understood the contents of the text.

\section{Sample source code}

All the source code of the programs that I wrote during this project
can be found at the following address:

\begin{center}
\url{https://github.com/Erkaman/digital-image-formats}.
\end{center}

\noindent The project is divided into several directories. In the \path|code|
directory the \CC source files for the programs written during the
project can be found. The \path|log| directory contains the log book
of this project. And in the \path|latex| directory you can find the
\LaTeX{} source files that were compiled to produce this document. But
even if you do not understand anything of the contents of the
\path|code| folder, do not worry, because this text was written to
explain the concepts behind image formats as intuitively as possible,
without you having to understand the source files for the programs
that demonstrates them. But it is of course not a bad thing if you do
understand the source code files in the \path|code| directory.
