\begin{comment}
  \bibliography{project.bib}
\end{comment}

\chapter{Purpose and problem}

\section{Purpose}

So the main purpose of this project is to explore how image formats
works. But only the area of image formats itself is a very broad area
since there are \textit{so many} of them. Finding out exactly how
many image formats there are in existence is a very difficult, almost
impossible, task. Imagemagick, an open source project whose main purpose
could be said to be to support the maximum amount of image formats,
supports well over 100 formats \cite{11:imagemagick_home}. And
according to the open directory project there are at this time about
170 different graphics formats \cite{10:opendirectory_data_formats}.

Indeed there are many different kinds of image formats. And
furthermore, image formats are generally divided into two main
categories: \textit{vector and raster
  graphics}\cite{murray1996encyclopedia,roelofs99:_png}. Vector
graphics consists of only lines, rectangles, polygons and other
geometric forms, while raster graphics could be described as a grid,
matrix, of colors. Neither of these is really simpler than the other,
but in this project I will focus only on raster graphics. Some
authors, like \cite{murray1996encyclopedia}, also refer raster
graphics to as \textit{bitmap images}.

There are also many different raster formats. Yet all of them are very
similar in their overall structure. So to make the project even more
manageable, I will only cover the three specific raster formats that
are known as TGA, GIF and PNG. The simplest formats will be covered
first, and the most the complex format will be covered last, meaning
that I will explore TGA, then GIF and last PNG, which is significantly
more complex than the former two.

By exploring these 3 formats, I will have gotten a good general view
of how image formats are built and structured. So this is main purpose
of the project:

\begin{quote}
  To explore how images are \textit{numerically} represented and
  structured digitally.
\end{quote}

And by this I mean that this project will go on until I have fully
understood the specifications of the 3 image formats TGA, GIF and PNG,
which by no means is as easy as it sounds, considering how terse these
documents are. By understanding the specification of an image format
you have fully understood that image format. Then you know about and
understand every single image of that format.
